Effects of epidemiological structure on the transient evolution of HIV virulence

### 2016-10-03 ms/plos_latex.tex
update to the manuscript (still need more updates)

%% Include all macros below
\newcommand{\khalf}{\left(\frac{1}{2}\right)^{\delta_{ij}}}  % (1/2)^kronecker
\newcommand{\Lspvl}{$\log_{10}$ SPVL}
\newcommand{\rzero}{{\mathcal R}_0}
\newcommand{\etal}{\textit{et al.}}
\newcommand{\tsub}[2]{#1_{{\textrm{\tiny #2}}}}
\newcommand{\PF}{\textrm{PF}}
\newcommand{\DS}{\textrm{DS}}
\newcommand{\WT}{\textrm{WT}}
\newcommand{\ET}{\textrm{ET}}
\newcommand{\DM}{\textrm{DM}}

@@

We developed six multi-strain evolutionary models covering a gamut between Champredon \etal's relatively realistic \cite{champredon_hiv_2013} and Shirreff \etal's relatively simple \cite{shirreff_transmission_2011} epidemiological structures, each of which is based on different assumptions regarding contact structure and partnership dynamics. Specifically, we focus on the effects of the assumptions of (1) instantaneous vs. non-instantaneous partnership formation and (2) zero vs. positive extra-partnership sexual contact and transmission on the evolution of mean \Lspvl.

@@

The last two models do not explicitly track sexual partnerships.  One (``implicit'') is an implicit serial monogamy model based on the epidemiological model used by Shirreff \etal\ \cite{shirreff_transmission_2011}. It is actually a random mixing model that consist of only two states, $S$ and $I$, and does not consider explicit partnership dynamics. However, to reflect the effect of (instantaneously formed) partnership structure, it uses an adjusted transmission rate that is derived from an approximation of the basic reproduction number of a serial monogamy model \cite{hollingsworth_hiv1_2008}. Finally, the last model (``random'') is a simple random-mixing model.

The base model (i.e. pairform+epc) for the first four models is an extension of Champredon \etal's model \cite{champredon_hiv_2013}. Individuals in single compartment acquire a partner at a rate $\rho$, and partnerships dissolve at a rate $c$. Infected individuals in a discordant partnership infect their susceptible partner at a rate $\beta$ (within-couple transmission rate) and susceptible individuals outside the partnership at a rate $c_e$ (extra-couple transmission rate). Likewise, a single infected individual can infect any susceptible individuals at a rate $c_u$ through uncoupled mixing. Extra-couple and uncoupled transmission are modeled in the same way as Champredon \etal's model. All the details have been adapted to a multi-strain scenario. The second through fourth models (pairform, instswitch+epc, instswitch) are derived from the base model by simplifying epidemiological processes (partnership formation and uncoupled/extra-couple contact). Model details are explained in the appendix.

@@

Despite considerable effort \cite{hollingsworth_hiv1_2008,champredon_hiv_2013}, the parameters determining the rate and structure of sexual partnership change and contact are still very uncertain; this led Champredon \etal\ \cite{champredon_hiv_2013} to adopt a Latin hypercube sampling (LHS) strategy \cite{blower_drugs_1991} that evaluates model outcomes over a range of parameter values. In order to make sure that our comparisons among models apply across the entire space of reasonable parameter values, and in order to evaluate the differential sensitivity of different model structures to parameter values, we follow a similar protocol and perform LHS over a parameter set including both the early- and late-stage transmission and duration parameters ($\beta_P$, $D_P$, $\beta_D$, $D_D$) and contact/partnership parameters ($\rho$, $c$, $c_u/c_w$, and $c_e/c_w$). We do not allow for uncertainties in parameters that are directly related to the evolutionary process ($\tsub{\beta}{max}$, $\beta_{50}$, $\beta_k$, $\tsub{D}{max}$, $D_{50}$, $D_k$, $\sigma_M$), using Shirreff \etal's point estimates throughout \cite{shirreff_transmission_2011}.

@@

The calibration runs for each parameter set are slightly
simplified. We run each model for only 500 years
(full simulations are run for 4000 years), and we use a 4/5 order 
Runge-Kutta method (\texttt{ode45} from the \texttt{deSolve} package
\cite{soetaert_solving_2010}), whereas we use
a stiff solver (LSODA) for the full simulations.

@@

Single individuals acquire partners at per-person rate $\rho$. Partnership formation rate for $S$, $I_i$ are $- \rho S$ and $-\rho I_i$, respectively. We follow Champredon \etal's results \cite{champredon_hiv_2013} and assume that single individuals are distributed into coupled states with pair-formation (PF) rates as follows:

@@

Within-couple transmission (WT) occurs in both models. An infected partner in $SI$ partnership transmits virus to a susceptible partner, and partnership state becomes $II$: $\WT(SI_i) = - \beta_i SI_i$. Since we assume that mutation occurs, $II_{ij}$, where $i \neq j$, can be formed from both $SI_i$ and $SI_j$ partnership: $\WT(II_{ij}) = M_{ij} \beta_i SI_i + M_{ji} \beta_j SI_j$. On the other hand, $II_{ii}$ can only be formed from an $SI_i$ partnership: $\WT(II_{ii}) = M_{ii} \beta_i SI_i$. Using the Kronecker delta notation, we obtain the following set of equations that describes within-couple transmission dynamics:

@@

Single susceptible individuals become infected and enter single infected compartment at the total rate of $\sum_k P_k U_k S$. Through mutation, newly infected individuals are distributed into single infected compartments with different strains: $\ET(I_i) = \sum_k M_{ki} P_k U_k S$. Either partner in an $SS$ partnership can be infected, with the partnership state becoming $SI$, at the total rate of $\sum_i 2 P_i E_i SS$. The formation of $SI_i$ partnerships is similar to the process through which single susceptible individuals are distributed into single infected compartments: $\ET(SI_i) = \sum_k 2 M_{ki} P_k E_k SS$. Lastly, the susceptible partner in an $SI$ partnership can be infected due to uncoupled/extra-couple contacts and partnership can change to an $II$ partnership. As in the previous cases, $SI_i$ partnerships are lost at a rate of $\sum_k P_k E_k SI_i$. The mutation process is similar to that of within-couple transmission. The only difference is that the \Lspvl\ of a newly infected partner is not determined by its social partner but from an extra-couple partner (i.e. the term $P_i$): $\ET(II_{ij}) = (\frac{1}{2})^{\delta_{ij}}(\sum_k (M_{kj} P_k E_k SI_i + M_{ki} P_k E_k SI_j))$. Combining these equations we get the following set of equations that describe all the transmission dynamics:

@@

>>>>

%% Include all macros below
\newcommand{\khalf}{\left(\frac{1}{2}\right)^{\delta_{ij}}}  % (1/2)^kronecker
\newcommand{\kkhalf}{\left(\frac{1}{2}\right)^{\delta_{ij} \delta_{kl}}}  % (1/2)^(kronecker * kronecker)
\newcommand{\Lspvl}{$\log_{10}$ SPVL}
\newcommand{\rzero}{{\mathcal R}_0}
\newcommand{\etal}{\textit{et al.}}
\newcommand{\tsub}[2]{#1_{{\textrm{\tiny #2}}}}
\newcommand{\PF}{\textrm{PF}}
\newcommand{\DS}{\textrm{DS}}
\newcommand{\WT}{\textrm{WT}}
\newcommand{\ET}{\textrm{ET}}
\newcommand{\DM}{\textrm{DM}}

@@

We developed seven multi-strain evolutionary models covering a gamut between Champredon \etal's relatively realistic \cite{champredon_hiv_2013} and Shirreff \etal's relatively simple \cite{shirreff_transmission_2011} epidemiological structures, each of which is based on different assumptions regarding contact structure and partnership dynamics. Specifically, we focus on the effects of the assumptions of (1) instantaneous vs. non-instantaneous partnership formation, (2) zero vs. positive extra-partnership sexual contact and transmission and (3) homogeneous vs. heterogeneous mixing on the evolution of mean \Lspvl.

@@

The next two models do not explicitly track sexual partnerships.  One (``implicit'') is an implicit serial monogamy model based on the epidemiological model used by Shirreff \etal\ \cite{shirreff_transmission_2011}. It is actually a random mixing model that consist of only two states, $S$ and $I$, and does not consider explicit partnership dynamics. However, to reflect the effect of (instantaneously formed) partnership structure, it uses an adjusted transmission rate that is derived from an approximation of the basic reproduction number of a serial monogamy model \cite{hollingsworth_hiv1_2008}. Finally, the last model (``random'') is a simple random-mixing model.

Lastly, we add heterogeneous mixing to pairform+epc model. Individuals are divided into different risk groups based on the sexual activity level, and we assume that sexual activity level is directly proportional to number of non-cohabiting (extra-couple and uncoupled) partners per year [CITE]. Model details can be found in the appendix.

The base model (i.e. pairform+epc) for the first four models and the heterogeneous model is an extension of Champredon \etal's model \cite{champredon_hiv_2013}. Individuals in single compartment acquire a partner at a rate $\rho$, and partnerships dissolve at a rate $c$. Infected individuals in a discordant partnership infect their susceptible partner at a rate $\beta$ (within-couple transmission rate) and susceptible individuals outside the partnership at a rate $c_e$ (extra-couple transmission rate). Likewise, a single infected individual can infect any susceptible individuals at a rate $c_u$ through uncoupled mixing. Extra-couple and uncoupled transmission are modeled in the same way as Champredon \etal's model. All the details have been adapted to a multi-strain scenario. The second through fourth models (pairform, instswitch+epc, instswitch) are derived from the base model by simplifying epidemiological processes (partnership formation and uncoupled/extra-couple contact). Model details are explained in the appendix.

@@

Despite considerable effort \cite{hollingsworth_hiv1_2008,champredon_hiv_2013}, the parameters determining the rate and structure of sexual partnership change and contact are still very uncertain; this led Champredon \etal\ \cite{champredon_hiv_2013} to adopt a Latin hypercube sampling (LHS) strategy \cite{blower_drugs_1991} that evaluates model outcomes over a range of parameter values. In order to make sure that our comparisons among models apply across the entire space of reasonable parameter values, and in order to evaluate the differential sensitivity of different model structures to parameter values, we follow a similar protocol and perform LHS over a parameter set including both the early- and late-stage transmission and duration parameters ($\beta_P$, $D_P$, $\beta_D$, $D_D$) and contact/partnership parameters ($\rho$, $c$, $c_u/c_w$, and $c_e/c_w$). For the heterogeneity model, mean and coefficient of variance (CV) for number of non-cohabiting partners are sampled as well. We do not allow for uncertainties in parameters that are directly related to the evolutionary process ($\tsub{\beta}{max}$, $\beta_{50}$, $\beta_k$, $\tsub{D}{max}$, $D_{50}$, $D_k$, $\sigma_M$), using Shirreff \etal's point estimates throughout \cite{shirreff_transmission_2011}.

@@

The calibration runs for each parameter set are slightly
simplified. We run each model for only 500 years
(full simulations are run for 4000 years). We use a 4/5 order 
Runge-Kutta method (\texttt{ode45} from the \texttt{deSolve} package
\cite{soetaert_solving_2010}) for all simulations.

@@

Single individuals acquire partners at per-person rate $\rho$. Partnership formation rate for $S$, $I_i$ are $\rho S$ and $\rho I_i$, respectively. We follow Champredon \etal's results \cite{champredon_hiv_2013} and assume that single individuals are distributed into coupled states with pair-formation (PF) rates as follows:

@@

Within-couple transmission (WT) occurs in both models. An infected partner in $SI$ partnership transmits virus to a susceptible partner at per-partnership rate $\beta$: $\WT(SI_i) = - \beta_i SI_i$. Since we assume that mutation occurs, $II_{ij}$, where $i \neq j$, can be formed from both $SI_i$ and $SI_j$ partnership: $\WT(II_{ij}) = M_{ij} \beta_i SI_i + M_{ji} \beta_j SI_j$. On the other hand, $II_{ii}$ can only be formed from an $SI_i$ partnership: $\WT(II_{ii}) = M_{ii} \beta_i SI_i$. Using the Kronecker delta notation, we obtain the following set of equations that describes within-couple transmission dynamics:

@@

Single susceptible individuals become infected through uncoupled contact at per-person rate $\sum_k P_k U_k$ and enter single infected state. Through mutation, newly infected individuals are distributed into single infected compartments with different strains: $\ET(I_i) = \sum_k M_{ki} P_k U_k S$. Either partner in an $SS$ partnership becomes infected at per-person rate $\sum_k P_k E_k$, and partnership state changes to an $SI$ partnership at the total rate of $\sum_i 2 P_i E_i SS$. The formation of $SI_i$ partnerships is similar to the process through which single susceptible individuals are distributed into single infected compartments: $\ET(SI_i) = \sum_k 2 M_{ki} P_k E_k SS$. Lastly, the susceptible partner in an $SI$ partnership becomes infected from extra-couple contacts at a per-person rate of $\sum_k P_k E_k$, and partnership state changes to an $II$ partnership. As in the previous cases, $SI_i$ partnerships are lost at a rate of $\sum_k P_k E_k SI_i$. The mutation process is similar to that of within-couple transmission. The only difference is that the \Lspvl\ of a newly infected partner is not determined by its social partner but from an extra-couple partner (i.e. the term $P_i$): $\ET(II_{ij}) = (\frac{1}{2})^{\delta_{ij}}(\sum_k (M_{kj} P_k E_k SI_i + M_{ki} P_k E_k SI_j))$. Combining these equations we get the following set of equations that describe all the transmission dynamics:

@@

\subsection*{Heteregenous model}

We extend "pairform+epc model" by allowing for heterogeneity in sexual behaviour. Since pairform+epc model captures four distinct sexual behaviours -- pair formation, pair dissolution, extra-couple mixing, and uncoupled mixing -- we assume that all four parameters that model the mentioned behaviours are scaled by the same factor based on the risk group. In other words, an individual in a higher risk is more likely to form a stable partnership, leave a stable partnership, and engage in a extra-couple/uncoupled mixing. We denote this scaling parameter as $\varphi_i$ where $i$ is the risk group. For simplicity, we assume that the transmission rate per partnership is not affected by sexual behaviour.

Individuals in a risk group $i$ leave single state at a rate of $\varphi_i \rho$. Let $XY_{ij,kl}$ be a coupled state where $X$ and $Y$ are the infection status (susceptible or infected) of each partner, $k$ and $l$ are the risk groups $X$ and $Y$ belong to respectively, and $i$ and $j$ are the strains of an infected partner. If a partner is susceptible, strain index is replaced by $\cdot$. For example, $SI_{\cdot j,kl}$ is the number of partners where the susceptible partner is in risk group $k$ and infected partner is in risk group $l$ and has \Lspvl\ of $\alpha_j$. For simplicity, we assume that people mix randomly. Then, we can write the partnership formation process as follows:

\begin{equation}
\PF(XY_{ij, kl}) = \kkhalf \frac{\varphi_k \rho X_{i,k} \varphi_l \rho Y_{j, l}}{\sum\limits_m \varphi_m \rho (S_{\cdot, m} + \sum\limits_n I_{n,m})}
\end{equation}

For dissolution process, individual in risk group $i$ leave partnership at a rate $\varphi_i c$. If partnership is formed between two individuals from a different risk group, rate at which they leave the partnership differs. We resolve this conflict by assuming that a partnership dissolution rate of a couple is equal to the average of that of two partners. Therefore, $XY_{ij, kl}$ dissolve at a rate $\frac{\varphi_k + \varphi_l}{2} c XY_{ij, kl}$, and both $X_{i,k}$ and $Y_{j,l}$ return to single state at a rate of $\frac{\varphi_k + \varphi_l}{2} c XY_{ij, kl}$.

\subsubsection*{Heterogeneous models: Infection dynamics}

Since we assume that the rate of transmission per partnership stays constant across different risk groups, within-couple infection process is similar to other models:

\begin{equation}
\begin{aligned}
WT(SI_{\cdot j, kl} &= - \beta_j SI_{\cdot j, kl}\\
WT(II_{ij, kl}) &= \kkhalf \cdot (M_{ji} \beta_j SI_{\cdot j, kl} + M_{ij} \beta_i SI_{\cdot i, lk})
\end{aligned}
\end{equation}

Note that $II_{ij,kl}$ can be formed from two types of partnerships: 1) Infected partner with \Lspvl\ of $\alpha_j$ and risk group of $l$ infects a susceptible partner in risk group $k$, yielding \Lspvl\ of $\alpha_i$ through mutation. 2) Infected partner with \Lspvl\ of $\alpha_i$ and risk group of $k$ infected a susceptible partner in risk group $l$, yielding \Lspvl\ of $\alpha_j$ through mutation. On the other hand, if $i = j$ and $k = l$, $II_{ii,kk}$ can only be formed from $SI_{\cdot i, kk}$ partnership, which is resolved by $\kkhalf$.

Heterogeneous extra-couple and uncoupled contact process is similar to partnership formation process. Relative uncoupled/extra-couple contact rates are scaled by the factor of $\varphi_i$, where $i$ is the risk group. First, we define $Q_{i}$ as the total rate of uncoupled/extra couple contact by individuals in risk group $k$:

\begin{equation}
\begin{aligned}
Q_i = &\varphi_i r_u (S_{\cdot,i} +  \sum_j I_{j,i}) + \varphi_i r_e \bigg( \sum_k 2^{\delta_{ik}} SS_{\cdot, ik} +\\
&\sum_l \sum_j (SI_{\cdot j,il} + SI_{\cdot j, li}) + \sum_j \sum_l \sum_k 2^{\delta_{kl} \delta_{ij}} II_{kl,ij} \bigg)
\end{aligned}
\end{equation}
We now define $P_{k,i}$ as the proportion of the extra-couple and uncoupled contact that arises from an infected individual from risk group $i$ with \Lspvl\ of $\alpha_k$:
\begin{equation}
P_{k,i} = \frac{\varphi_i r_u I_{k,i} + \varphi_i r_e (SI_{k,i} + \sum_j \sum_l 2^{\delta_{kl} \delta_{ij} } II_{kl,ij} )}{\sum_j Q_j}
\end{equation}
Since the relative uncoupled/extra couple contact ratio are scaled by the factor of $\varphi_i$, uncoupled and extra-couple transmission rates are scaled by the same factor as well: $U_{k,i} = \varphi_i r_u \beta_k$ and $E_{k,i} = \varphi_i r_e \beta_k$. Once again, we assume random mixing between individuals. Then, a susceptible individual in risk group $i$ becomes infected through extra-couple and uncoupled contact at a per capita rate of $\sum_j \sum_k P_{k,j} X_{k,i}$. Once infected, individuals are distributed into each $\Lspvl$ categories through mutation.

\subsubsection*{Heterogeneous model: Disease induced mortality}

Disease induced mortality is not affected by the sexual behaviour of an individual.
